Abstractions for Multi-Sorted Substitutions

### 2024-03-14 paper/paper.tex
Update.

  As we want our framework to support dependent types, we allow a type returned
  by the typing context to use all variables bound previously in the typing context.
  This is achieved by the \ACode{drop-∈} function, which cuts off all
  sorts from \ACode{S} that are at smaller or equal positions than a variable \ACode{x}.

@@

  \section{Typing}
  \label{sec:typing}
  Now that we have a notion of types and type context, we are ready to define
  the multi-sorted typing relation for System F, incorporating both
  typing and kinding:
  \AppCode\FTyping

@@

  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of the lambda abstraction, which
    weakens the type of the codomain \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-kinded due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst

@@

  We then define a \ACode{\ACon{TTraversal}} analogously to
  \ACode{\ACon{Traversal}}, but instead of defining the application of
  maps, it defines the type preservation of applying maps:
  \LibCode\KTTraversal
  Given a term \ACode{e} with typing \ACode{⊢e} and
  a renaming/substitution \ACode{ϕ} with typing \ACode{⊢ϕ},
  the term \ACode{⊢e \AField{⊢⋯} ⊢ϕ} is a typing for \ACode{e \AField{⋯} ϕ}.
  The large amount of kit-parameters does not impose any restriction,
  as both our \ACode{\ACon{Kit}}s support the \ACode{\ACon{WkKit}}
  extension and can be composed arbitrarily. They only make the previously
  built machinery available to the proof of \ACode{\AField{\_⊢⋯\_}}.

@@

  We believe restriction 3 can be lifted partially to allow for
  example the encoding of linear types that requires context
  splitting.

@@

  \section{Conclusion}
  \label{sec:conclusion}
  We want to focus our future work in the following directions:
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}, allowing recursive
    occurences in container types like lists and finite functions, and
    adding support for substructural types as in
    \cite{DBLP:journals/corr/abs-2005-02247}.
  \item
    Extending the class of typing relations covered by
    the \ACode{\ACon{TTraversal}} abstraction and identifying
    a subclass for which it can be derived.
  \item
    Developing a decision procedure for equational
    substitution lemmas based on $\sigma$-calculus as in Autosubst
    \cite{DBLP:conf/itp/SchaferTS15, DBLP:conf/cpp/StarkSK19}.
  \item
    Developing abstractions and derived lemmas beyond substition and
    type preservation, e.g. for confluence or denotational semantics
    as in \cite{DBLP:journals/pacmpl/AllaisA0MM18}.
  \end{itemize}

>>>>

  As we want our framework to support dependent types, we allow a type returned
  by the typing context to use all variables bound previously in the typing context.
  This is achieved by the \ACode{drop-∈} function, which cuts off all
  sorts from \ACode{S} that are at smaller or equal positions than a variable \ACode{x}.
  Contexts \ACode{Γ} can be extended by a type \ACode{t} via \ACode{t \ADef{∷ᵗ} Γ}.

@@

  \section{Typing}
  \label{sec:typing}
  Now that we have a notion of types and type context, we are ready to define
  the multi-sorted typing relation for System F, which describes both
  typing and kinding:
  \AppCode\FTyping

@@

  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of a lambda abstraction, which
    weakens codomain type \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-kinded due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst

@@

  We then define a \ACode{\ACon{TTraversal}} analogously to
  \ACode{\ACon{Traversal}}, but instead of defining the application of
  % maps, it defines that the application of typed maps preserves term types:
  maps, it defines that the application of a typed map to a typed term yields a typed term:
  \LibCode\KTTraversal
  Given a term \ACode{e} with typing \ACode{⊢e} and
  a renaming/substitution \ACode{ϕ} with typing \ACode{⊢ϕ},
  the term \ACode{⊢e \AField{⊢⋯} ⊢ϕ} is a typing for \ACode{e \AField{⋯} ϕ}.
  The large amount of kit-parameters does not impose any restriction,
  as both our \ACode{\ACon{Kit}}s support the \ACode{\ACon{WkKit}}
  extension and can be composed arbitrarily. They only make the previously
  built machinery available to the proof of \ACode{\AField{\_⊢⋯\_}}.

@@

  % We believe restriction 3 can be lifted partially to allow for
  % example the encoding of linear types that requires context
  % splitting.

@@

  \section{Conclusion}
  \label{sec:conclusion}
  We have presented a framework for deriving untyped substitution and its properties, and
  and abstraction for proving type preservation of substitution with a single lemma.

  Using extrinsic typing allowed us to unify the renamings and
  substitutions for variables of arbitrary many syntactic categories.

  Furthermore, extrinsic typing allowed us treat untyped substitution
  and its properties separately from typing and type preservation.
  This has the benefit that the former can be derived automatically for virtually
  any object language, while the latter is more domain specific and
  can be handled independently.

  We believe our framework is a viable alternative to intrinsic
  typing, as the duplication is automated away.

  We believe our framework is particularly useful for polymorphic languages,
  where intrinsic typing requires necessarily to deal with multiple substitutions
  and their interactions. Furthermore, in more complex scenarios, intrinsic typing
  can easily lead to subst-hell, which is avoided with extrinsic typing.

  We belive the core limitation of our framework are the typing kits, which
  require a classical ternary typing relation. However, this problem is hard
  to avoid as typing relations can become arbitrarily complicated.
  To adapt the framework to more complicated type systems, only the typing
  kits and typing traversals need to be extended, which we have successfully
  done for linear typing ala Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} and
  general substructural typing ala Licata et al\cite{DBLP:conf/rta/LicataSR17}.
